Make any stupid plot you want

### 2021-05-10 draft/sections/math.tex
math

\section{Topological Equivariant Artist Model}
\label{sec:math}
To guide the implementation of structure preserving visualization components, we develop a mathematical formalism of visualization that specifies how these components preserve \textit{continuity} and \textit{equivariance}. Inspired by the analogous classes in Matplotlib\cite{hunterArchitectureOpenSource}, we call the transformation from data space to graphic space that these building block components implement the \textit{artist}.
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
The \textit{artist} \(\mathscr{\vartist}\) is a map from data \(\mathscr{\dtotal}\) to graphic \(\mathscr{\gtotal}\) fiber bundles. To explain how the \textit{artist} is a structure preserving map from data to graphic, we first model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as topological structures that encapsulate component types and continuity. We then discuss the functional maps from graphic to data (\autoref{sec:math:graphic:base}), data components to visual components (\autoref{sec:math:artist:nu}), and visual components into graphic (\autoref{sec:math:artist:q}) that make up the artist.

\subsection{Data Space \dtotal}
\label{sec:math:data}
We use fiber bundles as the data model because they are inclusive enough to express all the types of structures of data described in \autoref{sec:intro:data:tools}. A fiber bundle is a tuple \((\dtotal,\,\dbase,\,\pi ,\,\dfiber)\) defined by the projection map \(\pi\)
\begin{equation}
    \label{eq:fiber_bundle}
    \begin{tikzcd}
        \dfiber \arrow[r, hook] & \dtotal \arrow[r, "\pi"] & \dbase
    \end{tikzcd}
\end{equation}
that binds the components of the data in \dfiber\ to the continuity of the data encoded in \dbase. Our use of fiber bundles builds on Butler's work proposing that fiber bundles should be the common data abstraction for visualization data\cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}. The fiber bundle models the properties of data component types \dfiber\ (\autoref{sec:math:data:fiber}), the continuity of records \dbase\ (\autoref{sec:math:data:base}), the collections of records (\autoref{sec:math:data:section}), and the space \dtotal\ of all possible datasets with these components and continuity. By definition fiber bundles are locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood \(U\) the total space is the cartesian product \(\dbase\times\dfiber\).

@@

\subsubsection{Sheafs}
\label{sec:math:data:sheaf}
Dynamic visualizations require evaluating sections on different subspaces of \dbase; this can be achieved using a mathematical structure, called a sheaf \(\mathcal{O}\), for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces.  On the fiber bundle \dtotal, we can describe a sheaf as the collection of local sections \(\iota^{*}\dsection\)
\begin{equation}
    \label{eq:sheaf}
    \begin{tikzcd}[column sep=huge]
        \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
        U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
    \end{tikzcd}
\end{equation}
which are sections of \dtotal\ pulled back over local neighborhood \(U \subset \dtotal\) via the inclusion map \(\iota:\dtotal\rightarrow U\). The collation of sections enabled by sheafs is necessary for navigation techniques such as pan and zoom\cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}.

\subsubsection{Applications}
Using fiber bundles as the data abstraction allows the model to describe widely used data containers without sacrificing the semantic structure embedded in each container. For example, the section can be any instance of a numpy array\cite{harris2020array} that stores an image, such as an image where the \dbase\ is a 2D continuous plane and the \dfiber\ is \((\reals^{3}, \reals, \reals)\). In this fiber, the \(\reals^3\) components encode color, and the other two components are the x and y positions of the sampled data in the image. The continuity of the image is implicitly encoded in the array as the index, so the position components encode the resolution. Instead of an image, the numpy array could also store a 2D discrete table. The fiber may not change, but the \dbase\ would now be 0D discrete points. These different choices in topology indicate, for example, what sorts of interpolation would be appropriate when visualizing the data. 
Labeled containers can also be described in this framework because of the schema like structure of the fiber. One such example is a pandas series which stores a labeled list, another is a  dataframe\cite{jeff_reback_2020_3715232} which has the structure of a relational table. A series could store the values of \(\dsection^{(1)}\) and a second series could be  \(\dsection^{(2)}\), while a dataframe would have multiple components and each data frame would be a unique section \dsection. The ability to encode complexity in continuity and components is particularly beneficial when working with  N dimensional labeled data containers. For example, an xarray\cite{hoyer2017xarray} data cube that stores precipitation would be a section of a fiber bundle with a \dbase\ that is a continuous volume and components \textit{(time, latitude, longitude, precipitation)}. This section does not need to resolve to values immediately and instead can be an instance of a distributed data container, such as a dask array \cite{rocklinDaskParallelComputation2015}.

@@

For a visualization component to preserve continuity, we propose that there must exist a structure preserving surjective map \(\vindex: \gbase\rightarrow\dbase\) from the data base space \dbase\ to the graphic base space \gbase. Formally, we require that \dbase\ be a deformation retract\cite{RetractionTopology2020} of \gbase\ such that \dbase\ and \gbase\ have the same homotopy, meaning there is a continuous map from \gbase\ to \dbase\cite{weissteinHomotopy}. The surjective map \(\vindex: \gbase \rightarrow \dbase\)
\begin{equation}
    \begin{tikzcd}
        \dtotal \arrow[d, "\pi"'] & \gtotal \arrow[d, "\pi"'] \\
        \dbase                   & \gbase \arrow[l, "\vindex"']
    \end{tikzcd}
    \label{eq:math:graphic:vindex}
\end{equation}
goes from region \(\gbasepoint \in \gbase_{\dbasepoint}\) to its associated point \(\dbasepoint \in \dbase\). This means that if \(\vindex(\gbasepoint) = \dbasepoint\), the record at \dbasepoint\ is copied over the region \gbasepoint\ such that \(\dsection(\dbasepoint)=\vindex^*\dsection(\gbasepoint)\) where \(\vindex^*\dsection(\gbasepoint)\)  is \dsection\ pulled back over \gbase. The map \vindex\ is part of the implementation of the artist \(\mathcal{\vartist}\) and therefore is not defined in terms of the data; instead it is how we specify the constraint that the type of the graphic \textit{continuity} must be able to map to the type of the data \textit{continuity}.

@@

\subsubsection{Composition of Artists: +}
\begin{figure}[H]
\begin{subfigure}{.5\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/combined_artist.png}
    \caption{}
    \label{fig:math:combined}    
\end{subfigure}
\begin{subfigure}{.5\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/exploding_artist.png}
    \caption{}
    \label{fig:math:seperate}
\end{subfigure}
\caption{ In \autoref{fig:math:combined}, these artists are composited before being added to the image. Disjoint union of \dtotal\ aligns the two time series with the x and y axis so all these elements use a shared coordinate system. A more complex composition dictates that the legend is connected to the \dtotal\ such that it must use the same color as the data it is identifying. None of this machinery exists in \autoref{fig:math:seperate}, therefore each artist is added to the page independent of the other elements.}
\label{fig:artist_plus}
\end{figure}
Visualizations generally consist of more than one artist, commonly having visual elements such as the plot and axis labels and maybe legends. To generate these composite images, we define addition operators and specify the constraints for compositing artists. Given the family of artists $(\dtotal_i: i\in I)$ that are rendered to the same image, the + operator 
\begin{equation}
+ \coloneqq \underset{i\in I}{\sqcup} \dtotal_{i}
\end{equation}
defines a simple composition of artists. For example, in \autoref{fig:math:combined} the data is joined via disjoint union; doing so aligns the components in \dfiber\ such the \vchannel\ to the same component in \vfiber\ targets the same coordinate system. In \autoref{fig:math:seperate}, these artists are all added to the image independently of the other and therefore there are no constraints on where they are placed in the image. When artists share a base space $\dbase_2 \hookrightarrow \dbase_1$, a composition operator can be defined such that the artists are acting on different components of the same section. This type of composition is important for visualizations where elements update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views\cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.

\subsubsection{Equivalance class of artists}
\label{sec:artist_equivalance}
\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/math/equivalent_artists.png}
   \caption{Each scatter plot is generated via a unique artist function $\vartist_{i}$, but they only differ in aesthetic styling. Therefore, these artists are all members of an equivalence class $\vartist_{i} \in \vartisteq$}
    \label{fig:math:artist:equivalence}
\end{figure}
Representational invariance, as defined by Kindlmann and Scheidegger, is the notion that visualizations are equivalent if changing the visual representation, such as colors or shapes, does not change the meaning of the visualization\cite{kindlmannAlgebraicProcessVisualization2014}. By defining a criteria for invariance, we can evaluate whether two artists generate the same type of graphic and compare artists across libraries. We propose that visualizations are invariant if they are generated by artists that are members of an equivalence class 
\begin{equation*}
\{\vartist \in \vartisteq: \vartist_{1} \equiv \vartist_{2}\}
\label{eq:math:artist:aprime}
\end{equation*}
For example, every scatter plot in \autoref{fig:math:artist:equivalence} is a scatter of the same datasets mapped to the \textit{x position} and \textit{y position} in the same way. The scatter plots only differ in the choice of constant visual literals, differing in color and marker shape. Each scatter is generated by an artist $\vartist_{i}$, and every scatter is generated by a member of the equivalence class $\vartist_{i} \in \vartisteq$. Since it is impractical to implement a new artist for every single graphic, the equivalence class provides a way to evaluate an implementation of a generalized artist. 
\end{document}

>>>>

\section{Topological Equivariant Artist Model}
\label{sec:math}
To guide the implementation of structure preserving visualization components, we develop a mathematical formalism of visualization that specifies how these components preserve \textit{continuity} and \textit{equivariance}. Inspired by the analogous classes in Matplotlib \cite{hunterArchitectureOpenSource}, we call the transformation from data space to graphic space that these building block components implement the \textit{artist}.
\begin{equation}
    \label{eq:artist}
    \mathscr{\vartist}: \mathscr{\dtotal} \rightarrow \mathscr{\gtotal}
\end{equation}
The \textit{artist} \(\mathscr{\vartist}\) is a map from data \(\mathscr{\dtotal}\) to graphic \(\mathscr{\gtotal}\) fiber bundles. To explain how the \textit{artist} is a structure preserving map from data to graphic, we first model data (\autoref{sec:math:data}) and graphics (\autoref{sec:math:graphic}) as topological structures that encapsulate component types and continuity. We then discuss the functional maps from graphic to data (\autoref{sec:math:graphic:base}), data components to visual components (\autoref{sec:math:artist:nu}), and visual components into graphic (\autoref{sec:math:artist:q}) that make up the artist.

\subsection{Data Space \dtotal}
\label{sec:math:data}
We use fiber bundles as the data model because they are inclusive enough to express all the types of structures of data described in \autoref{sec:intro:data:tools}. A fiber bundle is a tuple \((\dtotal,\,\dbase,\,\pi ,\,\dfiber)\) defined by the projection map \(\pi\)
\begin{equation}
    \label{eq:fiber_bundle}
    \begin{tikzcd}
        \dfiber \arrow[r, hook] & \dtotal \arrow[r, "\pi"] & \dbase
    \end{tikzcd}
\end{equation}
that binds the components of the data in \dfiber\ to the continuity of the data encoded in \dbase. Our use of fiber bundles builds on Butler's work proposing that fiber bundles should be the common data abstraction for visualization data \cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}. The fiber bundle models the properties of data component types \dfiber\ (\autoref{sec:math:data:fiber}), the continuity of records \dbase\ (\autoref{sec:math:data:base}), the collections of records (\autoref{sec:math:data:section}), and the space \dtotal\ of all possible datasets with these components and continuity. By definition fiber bundles are locally trivial \cite{spanier1989algebraic,LocallyTrivialFibre}, meaning that over a localized neighborhood \(U\) the total space is the cartesian product \(\dbase\times\dfiber\).

@@

\subsubsection{Sheafs}
\label{sec:math:data:sheaf}
Dynamic visualizations require evaluating sections on different subspaces of \dbase; this can be achieved using a mathematical structure, called a sheaf \(\mathcal{O}\), for defining collections of objects\cite{ghristElementaryAppliedTopology2014, ghristHomologicalAlgebraData2018,urbanikBriefIntroductionSchemes} on mathematical spaces.  On the fiber bundle \dtotal, we can describe a sheaf as the collection of local sections \(\iota^{*}\dsection\)
\begin{equation}
    \label{eq:sheaf}
    \begin{tikzcd}[column sep=huge]
        \iota^*\dtotal \arrow[d, "\pi"'] \arrow[r, "\iota^*", hook]             & \dtotal \arrow[d, "\pi"']                  \\
        U \arrow[r, "\iota", hook] \arrow[u, "\iota^{*}\dsection"', bend right] & \dbase \arrow[u, "\dsection"', bend right]
    \end{tikzcd}
\end{equation}
which are sections of \dtotal\ pulled back over local neighborhood \(U \subset \dtotal\) via the inclusion map \(\iota:\dtotal\rightarrow U\). The collation of sections enabled by sheafs is necessary for navigation techniques such as pan and zoom \cite{NekrasovskiEvaluationPanZoom2006} and dynamically updated visualizations such as sliding windows\cite{crouchDynamicGraphsSlidingwindow2013,chuTimeSeriesSegmentation1995}.

\subsubsection{Applications}
Using fiber bundles as the data abstraction allows the model to describe widely used data containers without sacrificing the semantic structure embedded in each container. For example, the section can be any instance of a numpy array \cite{harris2020array} that stores an image, such as an image where the \dbase\ is a 2D continuous plane and the \dfiber\ is \((\reals^{3}, \reals, \reals)\). In this fiber, the \(\reals^3\) components encode color, and the other two components are the x and y positions of the sampled data in the image. The continuity of the image is implicitly encoded in the array as the index, so the position components encode the resolution. Instead of an image, the numpy array could also store a 2D discrete table. The fiber may not change, but the \dbase\ would now be 0D discrete points. These different choices in topology indicate, for example, what sorts of interpolation would be appropriate when visualizing the data. 
Labeled containers can also be described in this framework because of the schema like structure of the fiber. One such example is a pandas series which stores a labeled list, another is a  dataframe \cite{jeff_reback_2020_3715232} which has the structure of a relational table. A series could store the values of \(\dsection^{(1)}\) and a second series could be  \(\dsection^{(2)}\), while a dataframe would have multiple components and each data frame would be a unique section \dsection. The ability to encode complexity in continuity and components is particularly beneficial when working with  N d imensional labeled data containers. For example, an xarray \cite{hoyer2017xarray} data cube that stores precipitation would be a section of a fiber bundle with a \dbase\ that is a continuous volume and components \textit{(time, latitude, longitude, precipitation)}. This section does not need to resolve to values immediately and instead can be an instance of a distributed data container, such as a dask array \cite{rocklinDaskParallelComputation2015}.

@@

For a visualization component to preserve continuity, we propose that there must exist a structure preserving surjective map \(\vindex: \gbase\rightarrow\dbase\) from the data base space \dbase\ to the graphic base space \gbase. Formally, we require that \dbase\ be a deformation retract \cite{RetractionTopology2020} of \gbase\ such that \dbase\ and \gbase\ have the same homotopy, meaning there is a continuous map from \gbase\ to \dbase \cite{weissteinHomotopy}. The surjective map \(\vindex: \gbase \rightarrow \dbase\)
\begin{equation}
    \begin{tikzcd}
        \dtotal \arrow[d, "\pi"'] & \gtotal \arrow[d, "\pi"'] \\
        \dbase                   & \gbase \arrow[l, "\vindex"']
    \end{tikzcd}
    \label{eq:math:graphic:vindex}
\end{equation}
goes from region \(\gbasepoint \in \gbase_{\dbasepoint}\) to its associated point \(\dbasepoint \in \dbase\). This means that if \(\vindex(\gbasepoint) = \dbasepoint\), the record at \dbasepoint\ is copied over the region \gbasepoint\ such that \(\dsection(\dbasepoint)=\vindex^*\dsection(\gbasepoint)\) where \(\vindex^*\dsection(\gbasepoint)\)  is \dsection\ pulled back over \gbase. The map \vindex\ is part of the implementation of the artist \(\mathcal{\vartist}\) and therefore is not defined in terms of the data; instead it is how we specify the constraint that the type of the graphic \textit{continuity} must be able to map to the type of the data \textit{continuity}.

@@

\subsubsection{Composition of Artists: +}
\begin{figure}[H]
\begin{subfigure}{.5\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/combined_artist.png}
    \caption{}
    \label{fig:math:combined}    
\end{subfigure}
\begin{subfigure}{.5\textwidth}
    \includegraphics[width=1\textwidth]{figures/math/exploding_artist.png}
    \caption{}
    \label{fig:math:seperate}
\end{subfigure}
\caption{ In \autoref{fig:math:combined}, these artists are composited before being added to the image. Disjoint union of \dtotal\ aligns the two time series with the x and y axis so all these elements use a shared coordinate system. A more complex composition dictates that the legend is connected to the \dtotal\ such that it must use the same color as the data it is identifying. None of this machinery exists in \autoref{fig:math:seperate}, therefore each artist is added to the page independent of the other elements.}
\label{fig:artist_plus}
\end{figure}
Visualizations generally consist of more than one artist, commonly having visual elements such as the plot and axis labels and maybe legends. To generate these composite images, we define addition operators and specify the constraints for compositing artists. Given the family of artists $(\dtotal_i: i\in I)$ that are rendered to the same image, the + operator 
\begin{equation}
+ \coloneqq \underset{i\in I}{\sqcup} \dtotal_{i}
\end{equation}
defines a simple composition of artists. For example, in \autoref{fig:math:combined} the data is joined via disjoint union; doing so aligns the components in \dfiber\ such the \vchannel\ to the same component in \vfiber\ targets the same coordinate system. In \autoref{fig:math:seperate}, these artists are all added to the image independently of the other and therefore there are no constraints on where they are placed in the image. When artists share a base space $\dbase_2 \hookrightarrow \dbase_1$, a composition operator can be defined such that the artists are acting on different components of the same section. This type of composition is important for visualizations where elements update together in a consistent way, such as multiple views \cite{alboRadarComparativeEvaluation2016a, hullmanKeeping2018} and brush-linked views \cite{beckerBrushingScatterplots1987,bujaInteractiveData1991}.

\subsubsection{Equivalance class of artists}
\label{sec:artist_equivalance}
\begin{figure}[H]
    \includegraphics[width=1\textwidth]{figures/math/equivalent_artists.png}
   \caption{Each scatter plot is generated via a unique artist function $\vartist_{i}$, but they only differ in aesthetic styling. Therefore, these artists are all members of an equivalence class $\vartist_{i} \in \vartisteq$}
    \label{fig:math:artist:equivalence}
\end{figure}
Representational invariance, as defined by Kindlmann and Scheidegger, is the notion that visualizations are equivalent if changing the visual representation, such as colors or shapes, does not change the meaning of the visualization \cite{kindlmannAlgebraicProcessVisualization2014}. By defining a criteria for invariance, we can evaluate whether two artists generate the same type of graphic and compare artists across libraries. We propose that visualizations are invariant if they are generated by artists that are members of an equivalence class 
\begin{equation*}
\{\vartist \in \vartisteq: \vartist_{1} \equiv \vartist_{2}\}
\label{eq:math:artist:aprime}
\end{equation*}
For example, every scatter plot in \autoref{fig:math:artist:equivalence} is a scatter of the same datasets mapped to the \textit{x position} and \textit{y position} in the same way. The scatter plots only differ in the choice of constant visual literals, differing in color and marker shape. Each scatter is generated by an artist $\vartist_{i}$, and every scatter is generated by a member of the equivalence class $\vartist_{i} \in \vartisteq$. Since it is impractical to implement a new artist for every single graphic, the equivalence class provides a way to evaluate an implementation of a generalized artist. 
\end{document}
